---
title: "prepare_data"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
options(width = 100)
```

## Prepare Brazil Microregion Data

This script prepares the Brazil microregion dataset used by the downstream GLM, PCA, cluster analysis, and MANOVA workflows.

The final objects saved are:

- `df`: prepared analysis dataset
- `num_vars`: numeric variables used in downstream analyses

The output file is saved to:

```text
data/prepared_data.RData
```

```{r}
library(dplyr)
library(readr)
library(tibble)
library(rlang)
```

## Step 1: Define Paths

```{r}
# Raw CSV files are expected in the current project directory.
# Change input_dir if your raw CSV files are stored elsewhere.
input_dir <- "."

# Downstream scripts load /cloud/project/data/prepared_data.RData,
# so the relative output path should be data/prepared_data.RData.
output_dir <- "data"
output_file <- file.path(output_dir, "prepared_data.RData")

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

## Step 2: Helper Functions

```{r}
# Annual files: extract the 2019 column.
read_annual <- function(file, varname, year = "2019") {
  path <- file.path(input_dir, file)
  
  if (!file.exists(path)) {
    stop("File not found: ", path)
  }
  
  raw <- read_csv(path, show_col_types = FALSE)
  
  if (!year %in% names(raw)) {
    stop("Column '", year, "' was not found in ", file)
  }
  
  tibble(
    microregion_code = as.integer(raw[[1]]),
    !!varname := raw[[year]]
  )
}

# Weekly files: extract weeks beginning with 2019 and compute row mean or row sum.
read_weekly <- function(file, varname, year_prefix = "2019", use_sum = FALSE) {
  path <- file.path(input_dir, file)
  
  if (!file.exists(path)) {
    stop("File not found: ", path)
  }
  
  raw <- read_csv(path, show_col_types = FALSE)
  
  # Important: use the standard ASCII caret '^' for start-of-string matching.
  # This selects columns such as 201901, 201902, ..., 201952.
  week_cols <- grep(paste0("^", year_prefix), names(raw), value = TRUE)
  
  if (length(week_cols) == 0) {
    stop("No weekly columns beginning with '", year_prefix, "' were found in ", file)
  }
  
  x <- as.matrix(raw[, week_cols])
  values <- if (use_sum) {
    rowSums(x, na.rm = TRUE)
  } else {
    rowMeans(x, na.rm = TRUE)
  }
  
  tibble(
    microregion_code = as.integer(raw[[1]]),
    !!varname := values
  )
}

# Monthly file: for PDSI, extract 2019 month columns and compute row mean.
read_pdsi <- function(file, varname, month_pattern = "-19$") {
  path <- file.path(input_dir, file)
  
  if (!file.exists(path)) {
    stop("File not found: ", path)
  }
  
  raw <- read_csv(path, show_col_types = FALSE)
  month_cols <- grep(month_pattern, names(raw), value = TRUE)
  
  if (length(month_cols) == 0) {
    stop("No monthly columns matching '", month_pattern, "' were found in ", file)
  }
  
  x <- as.matrix(raw[, month_cols])
  values <- rowMeans(x, na.rm = TRUE)
  
  tibble(
    microregion_code = as.integer(raw[[1]]),
    !!varname := values
  )
}

# Regional median imputation for numeric variables.
impute_regional_median <- function(x) {
  med <- median(x, na.rm = TRUE)
  
  if (is.na(med)) {
    return(x)
  }
  
  ifelse(is.na(x), med, x)
}

# Winsorize a numeric vector at Q1 - k*IQR and Q3 + k*IQR.
winsorize_iqr <- function(x, k = 3) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lo <- q1 - k * iqr
  hi <- q3 + k * iqr
  pmax(pmin(x, hi), lo)
}
```

## Step 3: Load and Merge Source Files

```{r}
df <- list(
  read_annual("/cloud/project/data/Building_surface_area.csv", "building_area"),
  read_annual("/cloud/project/data/Gross_Domestic_Product.csv", "gdp"),
  read_weekly("/cloud/project/data/Mean_temperature.csv", "temp_mean"),
  read_weekly("/cloud/project/data/Minimum_temperature.csv", "temp_min"),
  read_weekly("/cloud/project/data/Maximum_temperature.csv", "temp_max"),
  read_weekly("/cloud/project/data/Mean_relative_humidity.csv", "humidity"),
  read_weekly("/cloud/project/data/Mean_atmospheric_pressure.csv", "pressure"),
  read_weekly("/cloud/project/data/Mean_wind_speed.csv", "wind_speed"),
  read_weekly("/cloud/project/data/Mean_Normalized_Difference_Vegetation_Index.csv", "ndvi"),
  read_weekly("/cloud/project/data/Sum_precipitation.csv", "precip", use_sum = TRUE),
  read_weekly("/cloud/project/data/Number_of_rainy_days.csv", "rainy_days", use_sum = TRUE),
  read_pdsi("/cloud/project/data/Palmer_Drought_Severity_Index.csv", "pdsi")
) %>%
  Reduce(function(x, y) inner_join(x, y, by = "microregion_code"), .)

dim(df)
head(df)
```

## Step 4: Add Region and Climate Variables

```{r}
state_to_region <- c(
  "11" = "Norte", "12" = "Norte", "13" = "Norte", "14" = "Norte",
  "15" = "Norte", "16" = "Norte", "17" = "Norte",
  "21" = "Nordeste", "22" = "Nordeste", "23" = "Nordeste",
  "24" = "Nordeste", "25" = "Nordeste", "26" = "Nordeste",
  "27" = "Nordeste", "28" = "Nordeste", "29" = "Nordeste",
  "31" = "Sudeste", "32" = "Sudeste", "33" = "Sudeste", "35" = "Sudeste",
  "41" = "Sul", "42" = "Sul", "43" = "Sul",
  "50" = "Centro-Oeste", "51" = "Centro-Oeste",
  "52" = "Centro-Oeste", "53" = "Centro-Oeste"
)

region_levels <- c("Centro-Oeste", "Nordeste", "Norte", "Sudeste", "Sul")
climate_levels <- c(
  "Tropical", "Semi-arid", "Tropical_Savanna", "Subtropical", "Temperate"
)

df <- df %>%
  mutate(
    state_code = microregion_code %/% 1000L,
    region = state_to_region[as.character(state_code)]
  )

if (any(is.na(df$region))) {
  missing_states <- sort(unique(df$state_code[is.na(df$region)]))
  stop("Some state codes were not mapped to regions: ", paste(missing_states, collapse = ", "))
}

df <- df %>%
  mutate(
    region = factor(region, levels = region_levels),
    climate_zone = case_when(
      region == "Norte" ~ "Tropical",
      region == "Nordeste" ~ "Semi-arid",
      region == "Centro-Oeste" ~ "Tropical_Savanna",
      region == "Sudeste" ~ "Subtropical",
      region == "Sul" ~ "Temperate",
      TRUE ~ NA_character_
    ),
    climate_zone = factor(climate_zone, levels = climate_levels)
  )

table(df$region)
table(df$climate_zone)
```

## Step 5: Define Numeric Variables

```{r}
num_vars <- c(
  "building_area", "gdp", "temp_mean", "temp_min", "temp_max",
  "humidity", "pressure", "wind_speed", "ndvi", "pdsi",
  "precip", "rainy_days"
)

missing_num_vars <- setdiff(num_vars, names(df))

if (length(missing_num_vars) > 0) {
  stop("The following numeric variables are missing: ", paste(missing_num_vars, collapse = ", "))
}

num_vars
```

## Step 6: Missing Values - Regional Median Imputation

```{r}
missing_before <- sum(is.na(df[, num_vars]))
cat("Missing numeric values before imputation:", missing_before, "\n")

df <- df %>%
  group_by(region) %>%
  mutate(across(all_of(num_vars), impute_regional_median)) %>%
  ungroup()

missing_after <- sum(is.na(df[, num_vars]))
cat("Missing numeric values after regional median imputation:", missing_after, "\n")
```

## Step 7: Create GDP Group

```{r}
# GDP groups are formed from the imputed GDP distribution before winsorization.
# This preserves the original tertile-based economic grouping while still allowing
# the continuous GDP variable to be winsorized for downstream numeric analyses.
gdp_breaks <- quantile(
  df$gdp,
  probs = c(0, 1/3, 2/3, 1),
  na.rm = TRUE
)

if (length(unique(gdp_breaks)) < length(gdp_breaks)) {
  stop("GDP quantile breaks are not unique; GDP grouping cannot be created safely.")
}

df <- df %>%
  mutate(
    gdp_group = cut(
      gdp,
      breaks = gdp_breaks,
      labels = c("Low", "Medium", "High"),
      include.lowest = TRUE
    ),
    gdp_group = factor(gdp_group, levels = c("Low", "Medium", "High"))
  )

table(df$gdp_group)
gdp_breaks
```

## Step 8: Outliers - Winsorize Continuous Variables at 3 x IQR

```{r}
# Winsorization is applied to numeric variables after imputation and after defining
# the GDP tertile groups.
for (col in num_vars) {
  df[[col]] <- winsorize_iqr(df[[col]], k = 3)
}
```

## Step 9: Final Checks and Summary

```{r}
cat("Dataset ready:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Categorical variables: region, climate_zone, gdp_group\n")
cat("Continuous variables:\n")
print(num_vars)
cat("Total missing values in final df:", sum(is.na(df)), "\n")

if (nrow(df) != 558) {
  warning("Expected 558 microregions, but final df has ", nrow(df), " rows.")
}

if (any(is.na(df$gdp_group))) {
  warning("gdp_group contains missing values.")
}

if (!all(table(df$gdp_group) == 186)) {
  warning("GDP groups are not exactly balanced at 186 observations per group.")
}

summary(df[, c("region", "climate_zone", "gdp_group", num_vars)])
```

## Step 10: Save Prepared Data

```{r}
save(df, num_vars, file = output_file)
cat("Saved prepared data to:", normalizePath(output_file), "\n")
```

```{r}
library(tidyverse)

# Replace df2019 with the name of your prepared 2019 analytic dataset
df <- df2019

# Continuous variables used for descriptive distribution checks
continuous_vars <- c(
  "building_area", "gdp",
  "temp_mean", "temp_min", "temp_max",
  "humidity", "pressure", "wind_speed",
  "ndvi", "pdsi", "precip", "rainy_days"
)

# Check that all variable names exist in the dataset
missing_vars <- setdiff(continuous_vars, names(df))
if (length(missing_vars) > 0) {
  stop("These variables are missing from df: ", paste(missing_vars, collapse = ", "))
}

# Long format for faceted plots
df_long <- df %>%
  select(all_of(continuous_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

# Mean and median lines for each variable
summary_lines <- df_long %>%
  group_by(variable) %>%
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(mean_value, median_value),
    names_to = "statistic",
    values_to = "xintercept"
  ) %>%
  mutate(
    statistic = recode(
      statistic,
      mean_value = "Mean",
      median_value = "Median"
    )
  )
```

