---
title: "Cluster Analysis"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
options(width = 100)
```

## Cluster Analysis on Brazil Microregion Data

This analysis clusters Brazilian microregions using selected socioeconomic, climate, and environmental variables. Because the clustering variables include both `gdp` and `building_area`, the clusters should be interpreted as **economic-environmental clusters**, not as purely climate clusters.

The four-cluster k-means solution is used as the primary solution because it provides an interpretable economic-environmental classification and, in the current output, gives the highest average silhouette width among k = 2 to k = 10. `NbClust` is treated as an additional diagnostic rather than as the sole decision rule.

Load required libraries.

```{r}
library(cluster)
library(factoextra)
library(NbClust)
library(ggplot2)
```

Load the prepared data and define variables.

```{r}
load("/cloud/project/data/prepared_data.RData")

# Basic checks
stopifnot(exists("df"))
stopifnot(exists("num_vars"))

# Keep a row identifier so clusters can be traced back to original observations.
df$.row_id <- seq_len(nrow(df))

# Stable factor ordering for interpretation.
if ("region" %in% names(df)) {
  region_levels <- c("Centro-Oeste", "Nordeste", "Norte", "Sudeste", "Sul")
  if (all(region_levels %in% unique(df$region))) {
    df$region <- factor(df$region, levels = region_levels)
  } else {
    df$region <- factor(df$region)
  }
}

if ("gdp_group" %in% names(df)) {
  df$gdp_group <- factor(df$gdp_group, levels = c("Low", "Medium", "High"))
}

# Selected variables for clustering.
# These follow the original PCA/cluster workflow after removing temp_min and temp_max.
cluster_vars <- c(
  "building_area", "gdp", "temp_mean", "humidity", "pressure",
  "wind_speed", "ndvi", "pdsi", "precip", "rainy_days"
)

# Confirm all variables exist.
missing_cluster_vars <- setdiff(cluster_vars, names(df))
if (length(missing_cluster_vars) > 0) {
  stop("The following clustering variables are missing from df: ",
       paste(missing_cluster_vars, collapse = ", "))
}

# Use complete cases for clustering.
cluster_needed <- unique(c(".row_id", cluster_vars, "region", "gdp_group"))
cluster_needed <- cluster_needed[cluster_needed %in% names(df)]

df_cluster <- df[complete.cases(df[, cluster_vars]), cluster_needed]

# Confirm data size and grouping variables.
dim(df_cluster)
summary(df_cluster[, cluster_vars])

if ("region" %in% names(df_cluster)) {
  table(df_cluster$region)
}

if ("gdp_group" %in% names(df_cluster)) {
  table(df_cluster$gdp_group)
}
```

---

## Step 1: Standardize Clustering Variables

Clustering is performed on standardized variables so that variables measured on large scales, especially `building_area`, `gdp`, and `precip`, do not dominate the distance calculation purely because of their units.

```{r}
kdata <- scale(df_cluster[, cluster_vars])

# Check standardization.
round(colMeans(kdata), 6)
round(apply(kdata, 2, sd), 6)
```

---

## Step 2: Inspect Correlations Among Clustering Variables

```{r}
cor_mat <- cor(df_cluster[, cluster_vars], use = "pairwise.complete.obs")
round(cor_mat, 2)
```

```{r correlation-heatmap, fig.width=7, fig.height=6, out.width="90%"}
# Cleaner heatmap than base::heatmap so axis labels are not clipped in the PDF.
cor_long <- as.data.frame(as.table(cor_mat))
names(cor_long) <- c("var1", "var2", "correlation")

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 2.7) +
  coord_equal() +
  scale_fill_gradient2(limits = c(-1, 1)) +
  labs(
    title = "Correlation Heatmap of Clustering Variables",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

---

## Step 3: Hierarchical Clustering with Euclidean Distance and Ward.D2

Ward clustering is conventionally paired with Euclidean distance. We therefore use `dist(..., method = "euclidean")` and `hclust(..., method = "ward.D2")`.

```{r hierarchical-dendrogram, fig.width=7, fig.height=5, out.width="90%"}
dist_euclidean <- dist(kdata, method = "euclidean")

hc_ward <- hclust(dist_euclidean, method = "ward.D2")

plot(hc_ward,
     labels = FALSE,
     hang = -1,
     main = "Hierarchical Clustering: Euclidean Distance + Ward.D2",
     xlab = "Microregions",
     ylab = "Height")

rect.hclust(hc_ward, k = 4, border = 2:5)
```
```{r}
## Step 3b: Hierarchical Clustering with Complete Linkage (Comparison)

hc_complete <- hclust(dist_euclidean, method = "complete")

plot(hc_complete,
     labels = FALSE,
     hang = -1,
     main = "Hierarchical Clustering: Euclidean Distance + Complete Linkage",
     xlab = "Microregions",
     ylab = "Height")

rect.hclust(hc_complete, k = 4, border = 2:5)
```

---

## Step 4: Hierarchical Cluster Assignments for k = 4 and k = 6

This section keeps the hierarchical k = 4 and k = 6 assignments for comparison with k-means and with the earlier workflow. The primary clustering interpretation below is based on k-means k = 4.

```{r}
hc4 <- cutree(hc_ward, k = 4)
hc6 <- cutree(hc_ward, k = 6)

df_cluster$hc4 <- factor(hc4)
df_cluster$hc6 <- factor(hc6)

cat("Hierarchical k = 4 cluster sizes:\n")
print(table(df_cluster$hc4))

cat("Hierarchical k = 6 cluster sizes:\n")
print(table(df_cluster$hc6))

if ("region" %in% names(df_cluster)) {
  hc4_region_table <- table(Hierarchical_k4 = df_cluster$hc4,
                            Region = df_cluster$region)
  hc6_region_table <- table(Hierarchical_k6 = df_cluster$hc6,
                            Region = df_cluster$region)
  
  cat("Hierarchical k = 4 counts by region:\n")
  print(hc4_region_table)
  cat("Hierarchical k = 4 row proportions by region:\n")
  print(round(prop.table(hc4_region_table, margin = 1), 3))
  
  cat("Hierarchical k = 6 counts by region:\n")
  print(hc6_region_table)
  cat("Hierarchical k = 6 row proportions by region:\n")
  print(round(prop.table(hc6_region_table, margin = 1), 3))
}

if ("gdp_group" %in% names(df_cluster)) {
  hc4_gdp_table <- table(Hierarchical_k4 = df_cluster$hc4,
                         GDP_Group = df_cluster$gdp_group)
  hc6_gdp_table <- table(Hierarchical_k6 = df_cluster$hc6,
                         GDP_Group = df_cluster$gdp_group)
  
  cat("Hierarchical k = 4 counts by GDP group:\n")
  print(hc4_gdp_table)
  cat("Hierarchical k = 4 row proportions by GDP group:\n")
  print(round(prop.table(hc4_gdp_table, margin = 1), 3))
  
  cat("Hierarchical k = 6 counts by GDP group:\n")
  print(hc6_gdp_table)
  cat("Hierarchical k = 6 row proportions by GDP group:\n")
  print(round(prop.table(hc6_gdp_table, margin = 1), 3))
}
```

---

## Step 5: Evaluate Candidate Number of Clusters

We evaluate k = 2 to k = 10 using within-cluster sum of squares, average silhouette width, and `NbClust`. The final number of clusters is selected by combining diagnostics with interpretability.

```{r elbow-plot, fig.width=7, fig.height=5, out.width="90%"}
set.seed(20260430)

k_range <- 2:10

wss <- sapply(k_range, function(k) {
  km <- kmeans(kdata, centers = k, nstart = 50, iter.max = 100)
  km$tot.withinss
})

plot(k_range, wss,
     type = "b",
     pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow Plot for k-means")
```

```{r silhouette-evaluation-plot, fig.width=7, fig.height=5, out.width="90%"}
sil_width <- sapply(k_range, function(k) {
  km <- kmeans(kdata, centers = k, nstart = 50, iter.max = 100)
  sil <- silhouette(km$cluster, dist_euclidean)
  mean(sil[, "sil_width"])
})

plot(k_range, sil_width,
     type = "b",
     pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette Evaluation for k-means")
```

```{r}
cluster_eval <- data.frame(
  k = k_range,
  total_withinss = wss,
  avg_silhouette = sil_width
)
cluster_eval

best_silhouette_k <- cluster_eval$k[which.max(cluster_eval$avg_silhouette)]
cat("Best k by average silhouette width:", best_silhouette_k, "\n")
```

### NbClust Evaluation

```{r nbclust-evaluation, fig.width=7, fig.height=5, out.width="90%"}
set.seed(20260430)

nbclust_result <- tryCatch(
  NbClust::NbClust(
    data = as.data.frame(kdata),
    distance = "euclidean",
    min.nc = min(k_range),
    max.nc = max(k_range),
    method = "ward.D2",
    index = "all"
  ),
  error = function(e) {
    message("NbClust failed. Continuing without NbClust output. Error message: ", e$message)
    NULL
  }
)

nbclust_votes_valid <- NULL
nbclust_votes_raw <- NULL

if (!is.null(nbclust_result)) {
  nb_best <- nbclust_result$Best.nc
  nb_best
  
  if (is.matrix(nb_best)) {
    best_k_values_raw <- nb_best[1, ]
    best_k_values_raw <- best_k_values_raw[!is.na(best_k_values_raw)]
    
    # Some graphical indices can return 0. Those are not admissible cluster numbers
    # and should not be counted as valid votes for k.
    best_k_values_valid <- best_k_values_raw[best_k_values_raw %in% k_range]
    
    nbclust_votes_raw <- sort(table(best_k_values_raw), decreasing = TRUE)
    nbclust_votes_valid <- sort(table(best_k_values_valid), decreasing = TRUE)
    
    cat("NbClust raw vote summary, before filtering invalid values:\n")
    print(nbclust_votes_raw)
    
    cat("NbClust valid vote summary, restricted to k = 2 to 10:\n")
    print(nbclust_votes_valid)
  }
}
```

---

## Step 6: Primary k-means Solution with k = 4

The k = 4 solution is retained as the primary interpretable economic-environmental cluster solution. We use a fixed random seed, `nstart = 50`, and `iter.max = 100` to improve reproducibility.

```{r}
set.seed(20260430)

km4 <- kmeans(
  kdata,
  centers = 4,
  nstart = 50,
  iter.max = 100
)

df_cluster$km4 <- factor(km4$cluster)

cat("k-means k = 4 cluster sizes:\n")
print(table(df_cluster$km4))

cat("k-means k = 4 within-cluster SS by cluster:\n")
print(km4$withinss)

cat("k-means k = 4 total within-cluster SS:", round(km4$tot.withinss, 4), "\n")
cat("k-means k = 4 between-cluster SS / total SS:",
    round(km4$betweenss / km4$totss, 4), "\n")
cat("k-means k = 4 iterations:", km4$iter, "\n")
cat("k-means k = 4 ifault code:", km4$ifault, "\n")
```

---

## Step 7: Describe k = 4 Cluster Centers

Cluster centers are shown both in standardized units and original units. Standardized centers are useful for comparing variable profiles; original-unit centers are easier to interpret substantively.

```{r}
# Standardized centers.
km4_centers_z <- as.data.frame(km4$centers)
km4_centers_z$cluster <- factor(seq_len(nrow(km4_centers_z)))
km4_centers_z <- km4_centers_z[, c("cluster", cluster_vars)]
km4_centers_z

# Original-unit means by k-means cluster.
km4_centers_original <- aggregate(df_cluster[, cluster_vars],
                                  by = list(km4 = df_cluster$km4),
                                  FUN = mean)
km4_centers_original

# Standard deviations by k-means cluster in original units.
km4_sds_original <- aggregate(df_cluster[, cluster_vars],
                              by = list(km4 = df_cluster$km4),
                              FUN = sd)
km4_sds_original
```

---

## Step 8: k = 4 Cluster Composition by Region and GDP Group

```{r}
if ("region" %in% names(df_cluster)) {
  km4_region_table <- table(Cluster = df_cluster$km4,
                            Region = df_cluster$region)
  
  cat("k-means k = 4 counts by region:\n")
  print(km4_region_table)
  
  cat("k-means k = 4 row proportions by region:\n")
  km4_region_prop <- prop.table(km4_region_table, margin = 1)
  print(round(km4_region_prop, 3))
}

if ("gdp_group" %in% names(df_cluster)) {
  km4_gdp_table <- table(Cluster = df_cluster$km4,
                         GDP_Group = df_cluster$gdp_group)
  
  cat("k-means k = 4 counts by GDP group:\n")
  print(km4_gdp_table)
  
  cat("k-means k = 4 row proportions by GDP group:\n")
  km4_gdp_prop <- prop.table(km4_gdp_table, margin = 1)
  print(round(km4_gdp_prop, 3))
}
```

---

## Step 9: Visualize k = 4 Clusters in PCA Space

```{r pca-space-base, fig.width=7, fig.height=5, out.width="90%"}
pca_cluster <- prcomp(kdata, center = FALSE, scale. = FALSE)

pca_scores <- as.data.frame(pca_cluster$x[, 1:2])
pca_scores$km4 <- df_cluster$km4
pca_scores$region <- if ("region" %in% names(df_cluster)) df_cluster$region else NA
pca_scores$gdp_group <- if ("gdp_group" %in% names(df_cluster)) df_cluster$gdp_group else NA

var_explained <- summary(pca_cluster)$importance[2, 1:2]

plot(pca_scores$PC1, pca_scores$PC2,
     col = as.integer(pca_scores$km4),
     pch = 19,
     xlab = paste0("PC1 (", round(100 * var_explained[1], 1), "%)"),
     ylab = paste0("PC2 (", round(100 * var_explained[2], 1), "%)"),
     main = "k-means k = 4 Clusters in PCA Space")
legend("topright",
       legend = levels(pca_scores$km4),
       col = seq_along(levels(pca_scores$km4)),
       pch = 19,
       title = "Cluster")
```

```{r pca-space-factoextra, fig.width=7, fig.height=5, out.width="90%"}
factoextra::fviz_cluster(
  km4,
  data = kdata,
  geom = "point",
  ellipse.type = "convex",
  main = "k-means k = 4 Economic-Environmental Clusters"
)
```

---

## Step 10: Sensitivity Analyses for Alternative k Values

The current diagnostics show that k = 4 has the highest average silhouette width, while `NbClust` votes may support alternative k values. Therefore, we examine k = 2 and k = 5 as diagnostic sensitivity analyses. We also keep k = 6 as an optional continuity check because it was considered in the earlier workflow.

```{r}
run_kmeans_solution <- function(k, kdata, df_cluster, cluster_vars, dist_obj) {
  set.seed(20260430 + k)
  km <- kmeans(kdata, centers = k, nstart = 50, iter.max = 100)
  cluster_factor <- factor(km$cluster)
  sil <- silhouette(km$cluster, dist_obj)
  
  centers_z <- as.data.frame(km$centers)
  centers_z$cluster <- factor(seq_len(nrow(centers_z)))
  centers_z <- centers_z[, c("cluster", cluster_vars)]
  
  centers_original <- aggregate(df_cluster[, cluster_vars],
                                by = list(cluster = cluster_factor),
                                FUN = mean)
  
  list(
    k = k,
    km = km,
    cluster_factor = cluster_factor,
    avg_silhouette = mean(sil[, "sil_width"]),
    centers_z = centers_z,
    centers_original = centers_original,
    silhouette = sil
  )
}

sensitivity_ks <- c(2, 5, 6)
sensitivity_results <- lapply(sensitivity_ks, run_kmeans_solution,
                              kdata = kdata,
                              df_cluster = df_cluster,
                              cluster_vars = cluster_vars,
                              dist_obj = dist_euclidean)
names(sensitivity_results) <- paste0("k", sensitivity_ks)

for (name in names(sensitivity_results)) {
  res <- sensitivity_results[[name]]
  k <- res$k
  
  df_cluster[[paste0("km", k)]] <- res$cluster_factor
  
  cat("\n==============================\n")
  cat("Sensitivity k =", k, "\n")
  cat("==============================\n")
  
  cat("Cluster sizes:\n")
  print(table(res$cluster_factor))
  
  cat("Between-cluster SS / total SS:",
      round(res$km$betweenss / res$km$totss, 4), "\n")
  cat("Average silhouette:", round(res$avg_silhouette, 4), "\n")
  cat("Iterations:", res$km$iter, "\n")
  cat("ifault code:", res$km$ifault, "\n")
  
  cat("Standardized centers:\n")
  print(res$centers_z)
  
  cat("Original-unit centers:\n")
  print(res$centers_original)
  
  if ("region" %in% names(df_cluster)) {
    reg_tab <- table(Cluster = res$cluster_factor,
                     Region = df_cluster$region)
    cat("Counts by region:\n")
    print(reg_tab)
    cat("Row proportions by region:\n")
    print(round(prop.table(reg_tab, margin = 1), 3))
  }
  
  if ("gdp_group" %in% names(df_cluster)) {
    gdp_tab <- table(Cluster = res$cluster_factor,
                     GDP_Group = df_cluster$gdp_group)
    cat("Counts by GDP group:\n")
    print(gdp_tab)
    cat("Row proportions by GDP group:\n")
    print(round(prop.table(gdp_tab, margin = 1), 3))
  }
}
```

---

## Step 11: Compare Primary k = 4 with Alternative k Values

```{r}
for (k in sensitivity_ks) {
  cat("\nCross-tabulation: primary km4 vs sensitivity k =", k, "\n")
  print(table(km4 = df_cluster$km4,
              sensitivity = df_cluster[[paste0("km", k)]]))
}

sil4 <- silhouette(km4$cluster, dist_euclidean)

silhouette_summary <- data.frame(
  k = c(4, sensitivity_ks),
  avg_silhouette = c(
    mean(sil4[, "sil_width"]),
    sapply(sensitivity_results, function(x) x$avg_silhouette)
  ),
  between_ss_over_total_ss = c(
    km4$betweenss / km4$totss,
    sapply(sensitivity_results, function(x) x$km$betweenss / x$km$totss)
  )
)

silhouette_summary <- silhouette_summary[order(silhouette_summary$k), ]
silhouette_summary
```

```{r silhouette-primary-k4, fig.width=7, fig.height=5, out.width="90%"}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
plot(sil4, main = "Silhouette Plot: Primary k = 4")
```

```{r silhouette-sensitivity-k2, fig.width=7, fig.height=5, out.width="90%"}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
plot(sensitivity_results[["k2"]]$silhouette,
     main = "Silhouette Plot: Sensitivity k = 2")
```

```{r silhouette-sensitivity-k5, fig.width=7, fig.height=5, out.width="90%"}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
plot(sensitivity_results[["k5"]]$silhouette,
     main = "Silhouette Plot: Sensitivity k = 5")
```

```{r silhouette-sensitivity-k6, fig.width=7, fig.height=5, out.width="90%"}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
plot(sensitivity_results[["k6"]]$silhouette,
     main = "Silhouette Plot: Sensitivity k = 6")
```

---

## Step 12: Optional - Export Cluster Assignments

```{r}
cluster_cols <- c(".row_id", "km4", paste0("km", sensitivity_ks), "hc4", "hc6")
cluster_cols <- cluster_cols[cluster_cols %in% names(df_cluster)]

cluster_assignments <- df_cluster[, cluster_cols]

if ("region" %in% names(df_cluster)) {
  cluster_assignments$region <- df_cluster$region
}

if ("gdp_group" %in% names(df_cluster)) {
  cluster_assignments$gdp_group <- df_cluster$gdp_group
}

head(cluster_assignments)

# Uncomment if you want to save the assignments.
# write.csv(cluster_assignments, "cluster_assignments.csv", row.names = FALSE)
```

---

## Final Summary

```{r}
cat("Cluster variables:\n")
print(cluster_vars)

cat("\nPrimary clustering method: k-means with k = 4, nstart = 50, iter.max = 100\n")
cat("Primary interpretation: economic-environmental clusters\n")
cat("Primary k = 4 between-cluster SS / total SS:", round(km4$betweenss / km4$totss, 4), "\n")
cat("Primary k = 4 average silhouette:", round(mean(sil4[, "sil_width"]), 4), "\n")
cat("Best k by average silhouette width across k = 2 to 10:", best_silhouette_k, "\n")

cat("\nCandidate-k evaluation table:\n")
print(cluster_eval)

if (!is.null(nbclust_votes_valid)) {
  cat("\nNbClust valid vote summary, excluding invalid k values:\n")
  print(nbclust_votes_valid)
}

cat("\nSensitivity summary:\n")
print(silhouette_summary)
```