---
title: "PCA"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
options(width = 80)
```

## Principal Component Analysis on Brazil Microregion Data

This PCA uses selected socioeconomic, climatic, and environmental variables to characterize Brazilian microregions. Because the PCA variables include both `gdp` and `building_area`, the PCA should be interpreted as an **economic-environmental PCA**, not as a purely climate/environmental PCA.

The highly correlated temperature variables `temp_min` and `temp_max` are excluded from the PCA because `temp_mean` already captures the main temperature gradient and because the three temperature variables are strongly collinear.

Load required libraries.

```{r}
library(corrplot)
library(PerformanceAnalytics)

# Prevent rgl/X11 warnings when knitting on a server or Posit Cloud.
options(rgl.useNULL = TRUE)
library(heplots)
```

Load the prepared data.

```{r}
load("/cloud/project/data/prepared_data.RData")

# Basic checks
stopifnot(exists("df"))
stopifnot(exists("num_vars"))

# Keep a row identifier for tracing observations.
df$.row_id <- seq_len(nrow(df))

# Stable factor ordering for interpretation.
if ("region" %in% names(df)) {
  region_levels <- c(
    "Centro-Oeste", "Nordeste", "Norte", "Sudeste", "Sul"
  )
  
  if (all(region_levels %in% unique(df$region))) {
    df$region <- factor(df$region, levels = region_levels)
  } else {
    df$region <- factor(df$region)
  }
}

if ("gdp_group" %in% names(df)) {
  df$gdp_group <- factor(
    df$gdp_group,
    levels = c("Low", "Medium", "High")
  )
}

# Confirm numeric variables available for PCA-related inspection.
num_vars
```

---

## Step 1: Check Correlations Among All Numeric Variables

Before running PCA, we examine whether variables are sufficiently correlated to justify dimension reduction. This correlation matrix includes all numeric variables in `num_vars`, including `temp_min` and `temp_max`, so that the temperature redundancy can be documented.

```{r}
# Correlation matrix using all numeric variables.
cor_all <- cor(df[, num_vars], use = "pairwise.complete.obs")
round(cor_all, 2)

# Visual correlation plot ordered by hierarchical clustering.
par(oma = c(0, 0, 2, 0), mar = c(1, 1, 1, 1))

corrplot.mixed(
  cor_all,
  lower.col = "black",
  upper = "ellipse",
  tl.col = "black",
  number.cex = 0.65,
  order = "hclust",
  tl.pos = "lt",
  tl.cex = 0.7
)

title("Correlations for Variables", outer = TRUE)
```

---

## Step 2: Check Linearity and Pairwise Structure

```{r}
chart.Correlation(df[, num_vars], histogram = TRUE, pch = 19)
```

---

## Step 3: Define Final PCA Variable Set

```{r}
# Remove temp_min and temp_max from num_vars.
# temp_mean, temp_min, and temp_max are highly correlated with one another.
# Retaining all three adds redundancy without adding much independent information.
num_vars2 <- num_vars[!num_vars %in% c("temp_min", "temp_max")]

# Confirm final PCA variable set.
num_vars2
length(num_vars2)
```

Create the PCA dataset using complete cases.

```{r}
# Include metadata for plotting and interpretation.
metadata_vars <- c(".row_id", "microregion_code", "region", "gdp_group")
metadata_vars <- metadata_vars[metadata_vars %in% names(df)]

pca_needed <- unique(c(metadata_vars, num_vars2))
df_pca <- df[complete.cases(df[, num_vars2]), pca_needed]

# Confirm dimensions and summaries.
dim(df_pca)
summary(df_pca[, num_vars2])

if ("region" %in% names(df_pca)) {
  table(df_pca$region)
}

if ("gdp_group" %in% names(df_pca)) {
  table(df_pca$gdp_group)
}
```

---

## Step 4: Check Correlations in the Final PCA Variable Set

```{r}
cor_pca <- cor(df_pca[, num_vars2], use = "pairwise.complete.obs")
round(cor_pca, 2)

par(oma = c(0, 0, 2, 0), mar = c(1, 1, 1, 1))

corrplot.mixed(
  cor_pca,
  lower.col = "black",
  upper = "ellipse",
  tl.col = "black",
  number.cex = 0.7,
  order = "hclust",
  tl.pos = "lt",
  tl.cex = 0.75
)

title("Correlations for Final PCA Variables", outer = TRUE)
```

---

## Step 5: Check Multivariate Normality

PCA does not require strict multivariate normality for descriptive dimension reduction, but this plot is useful for identifying departures from multivariate normality and potential outlying observations.

```{r mvn-check, fig.width=7, fig.height=5, out.width="90%"}
cqplot(
  scale(df_pca[, num_vars2]),
  main = "Multivariate Normality Check for Final PCA Variables"
)
```

---

## Step 6: Run PCA

We use `prcomp()` with `center = TRUE` and `scale. = TRUE`, so all variables are standardized before PCA. This is necessary because the variables have different measurement units.

```{r}
# Custom summary function: shows eigenvalues instead of standard deviations.
summary.PCA.JDRS <- function(x) {
  sum_JDRS <- summary(x)$importance
  sum_JDRS[1, ] <- sum_JDRS[1, ]^2
  attr(sum_JDRS, "dimnames")[[1]][1] <- "Eigenvals (Variance)"
  sum_JDRS
}

# Run PCA on standardized variables.
pc1 <- prcomp(df_pca[, num_vars2], center = TRUE, scale. = TRUE)

# Eigenvalues and proportion of variance explained.
pca_summary <- round(summary.PCA.JDRS(pc1), 3)
pca_summary

# Loadings: variables driving each principal component.
pca_loadings <- round(pc1$rotation, 3)
pca_loadings
```

Create a clean PCA summary table.

```{r}
eigenvalues <- pc1$sdev^2
prop_var <- eigenvalues / sum(eigenvalues)
cum_var <- cumsum(prop_var)

pca_table <- data.frame(
  PC = paste0("PC", seq_along(eigenvalues)),
  eigenvalue = eigenvalues,
  proportion_variance = prop_var,
  cumulative_variance = cum_var,
  retain_kaiser = eigenvalues > 1
)

pca_table
```

---

## Step 7: Identify Main Loadings for Each Principal Component

```{r}
get_top_loadings <- function(rotation_matrix, pcs = 1:4, top_n = 5) {
  out <- lapply(pcs, function(pc_index) {
    pc_name <- colnames(rotation_matrix)[pc_index]
    loadings <- rotation_matrix[, pc_index]
    top_vars <- names(sort(abs(loadings), decreasing = TRUE))[1:top_n]
    
    data.frame(
      PC = pc_name,
      variable = top_vars,
      loading = loadings[top_vars],
      abs_loading = abs(loadings[top_vars]),
      row.names = NULL
    )
  })
  
  do.call(rbind, out)
}

top_loadings <- get_top_loadings(
  pc1$rotation,
  pcs = 1:min(4, ncol(pc1$rotation)),
  top_n = 5
)

top_loadings
```

---

## Step 8: Scree Plot

```{r scree-plot, fig.width=7, fig.height=5, out.width="90%"}
screeplot(
  pc1,
  type = "lines",
  col = "red",
  lwd = 2,
  pch = 19,
  cex = 1.2,
  main = "Scree Plot - Economic-Environmental PCA"
)

abline(h = 1, lty = 2)
```

---

## Step 9: Parallel Analysis / Threshold Plot

This self-contained parallel analysis avoids relying on externally sourced functions. Random datasets are generated by independently permuting each variable, preserving each variable's marginal distribution while breaking cross-variable correlations. PCs with observed eigenvalues above the random threshold are candidates for retention.

```{r parallel-analysis, fig.width=7, fig.height=5, out.width="90%"}
parallel_pca_threshold <- function(
  data,
  n_iter = 500,
  quantile_level = 0.95,
  seed = 20260430
) {
  set.seed(seed)
  
  x <- as.data.frame(scale(data))
  p <- ncol(x)
  
  observed_eigenvalues <- prcomp(
    x,
    center = FALSE,
    scale. = FALSE
  )$sdev^2
  
  random_eigenvalues <- matrix(
    NA_real_,
    nrow = n_iter,
    ncol = p
  )
  
  for (i in seq_len(n_iter)) {
    random_x <- as.data.frame(lapply(x, sample, replace = FALSE))
    random_eigenvalues[i, ] <- prcomp(
      random_x,
      center = FALSE,
      scale. = FALSE
    )$sdev^2
  }
  
  random_mean <- apply(random_eigenvalues, 2, mean)
  random_quantile <- apply(
    random_eigenvalues,
    2,
    quantile,
    probs = quantile_level
  )
  
  out <- data.frame(
    PC = paste0("PC", seq_len(p)),
    observed_eigenvalue = observed_eigenvalues,
    random_mean_eigenvalue = random_mean,
    random_quantile_eigenvalue = random_quantile,
    retain_mean_rule = observed_eigenvalues > random_mean,
    retain_quantile_rule = observed_eigenvalues > random_quantile
  )
  
  plot(
    seq_len(p),
    observed_eigenvalues,
    type = "b",
    pch = 19,
    ylim = range(c(observed_eigenvalues, random_mean, random_quantile)),
    xlab = "Principal Component",
    ylab = "Eigenvalue",
    main = "Parallel Analysis for PCA"
  )
  
  lines(seq_len(p), random_mean, type = "b", pch = 17, lty = 2)
  lines(seq_len(p), random_quantile, type = "b", pch = 15, lty = 3)
  abline(h = 1, lty = 4)
  
  legend_labels <- c(
    "Observed",
    "Random mean",
    paste0("Random ", quantile_level * 100, "%"),
    "Kaiser = 1"
  )
  
  legend(
    "topright",
    legend = legend_labels,
    pch = c(19, 17, 15, NA),
    lty = c(1, 2, 3, 4),
    bty = "n"
  )
  
  out
}

parallel_results <- parallel_pca_threshold(
  df_pca[, num_vars2],
  n_iter = 500,
  quantile_level = 0.95,
  seed = 20260430
)

parallel_results
```

---

## Step 10: Score Plot for PC1 vs PC2

```{r score-plot-ellipse, fig.width=7, fig.height=5, out.width="90%"}
draw_score_ellipse <- function(
  x,
  y,
  level = 0.95,
  n = 200,
  col = "red",
  lwd = 2
) {
  ok <- is.finite(x) & is.finite(y)
  xy <- cbind(x[ok], y[ok])
  
  mu <- colMeans(xy)
  S <- cov(xy)
  eig <- eigen(S)
  
  angles <- seq(0, 2 * pi, length.out = n)
  circle <- rbind(cos(angles), sin(angles))
  radius <- sqrt(qchisq(level, df = 2))
  
  ellipse <- t(
    mu + radius *
      eig$vectors %*%
      diag(sqrt(eig$values)) %*%
      circle
  )
  
  lines(ellipse[, 1], ellipse[, 2], col = col, lwd = lwd)
}

pcx <- 1
pcy <- 2

xlab_pc <- paste0("PC", pcx, " (", round(100 * prop_var[pcx], 1), "%)")
ylab_pc <- paste0("PC", pcy, " (", round(100 * prop_var[pcy], 1), "%)")

plot(
  pc1$x[, pcx],
  pc1$x[, pcy],
  pch = 19,
  cex = 0.6,
  col = "darkblue",
  xlab = xlab_pc,
  ylab = ylab_pc,
  main = "PC1 vs PC2 Score Plot"
)

draw_score_ellipse(pc1$x[, pcx], pc1$x[, pcy], level = 0.95)

# Label only the most extreme observations in PC1-PC2 space to avoid overplotting.
score_dist <- rowSums(scale(pc1$x[, c(pcx, pcy)])^2)
top_score_idx <- order(score_dist, decreasing = TRUE)[
  1:min(10, length(score_dist))
]

if ("microregion_code" %in% names(df_pca)) {
  text(
    pc1$x[top_score_idx, pcx],
    pc1$x[top_score_idx, pcy],
    labels = df_pca$microregion_code[top_score_idx],
    pos = 3,
    cex = 0.6
  )
} else {
  text(
    pc1$x[top_score_idx, pcx],
    pc1$x[top_score_idx, pcy],
    labels = df_pca$.row_id[top_score_idx],
    pos = 3,
    cex = 0.6
  )
}
```

---

## Step 11: Custom Biplot for PC1 vs PC2

The standard base-R `biplot()` labels all observations and can become unreadable with 558 microregions. The custom biplot below shows observations as points and variables as loading arrows.

```{r custom-biplot, fig.width=7, fig.height=5, out.width="90%"}
custom_pca_biplot <- function(pc, choices = c(1, 2), prop_var) {
  pcx <- choices[1]
  pcy <- choices[2]
  
  scores_x <- pc$x[, pcx]
  scores_y <- pc$x[, pcy]
  loadings_x <- pc$rotation[, pcx]
  loadings_y <- pc$rotation[, pcy]
  
  x_range <- diff(range(scores_x))
  y_range <- diff(range(scores_y))
  loading_x_range <- diff(range(loadings_x))
  loading_y_range <- diff(range(loadings_y))
  
  scale_factor <- 0.35 * min(
    x_range / loading_x_range,
    y_range / loading_y_range
  )
  
  arrow_x <- loadings_x * scale_factor
  arrow_y <- loadings_y * scale_factor
  
  plot(
    scores_x,
    scores_y,
    pch = 19,
    cex = 0.45,
    col = "gray60",
    xlab = paste0("PC", pcx, " (", round(100 * prop_var[pcx], 1), "%)"),
    ylab = paste0("PC", pcy, " (", round(100 * prop_var[pcy], 1), "%)"),
    main = paste0("PCA Biplot: PC", pcx, " vs PC", pcy)
  )
  
  abline(h = 0, v = 0, lty = 2, col = "gray40")
  arrows(0, 0, arrow_x, arrow_y, length = 0.08, col = "red", lwd = 1.5)
  
  text(
    arrow_x * 1.12,
    arrow_y * 1.12,
    labels = rownames(pc$rotation),
    col = "red",
    cex = 0.8
  )
}

custom_pca_biplot(pc1, choices = c(1, 2), prop_var = prop_var)
```

---

## Step 12: Score Plot Colored by Region

```{r score-by-region, fig.width=7, fig.height=5, out.width="90%"}
if ("region" %in% names(df_pca)) {
  region_factor <- as.factor(df_pca$region)
  
  plot(
    pc1$x[, 1],
    pc1$x[, 2],
    col = as.numeric(region_factor),
    pch = 19,
    cex = 0.7,
    xlab = paste0("PC1 (", round(100 * prop_var[1], 1), "%)"),
    ylab = paste0("PC2 (", round(100 * prop_var[2], 1), "%)"),
    main = "PC1 vs PC2 Colored by Region"
  )
  
  legend(
    "bottomright",
    legend = levels(region_factor),
    col = seq_along(levels(region_factor)),
    pch = 19,
    cex = 0.8,
    title = "Region"
  )
}
```

---

## Step 13: Score Plot Colored by GDP Group

```{r score-by-gdp-group, fig.width=7, fig.height=5, out.width="90%"}
if ("gdp_group" %in% names(df_pca)) {
  gdp_factor <- as.factor(df_pca$gdp_group)
  
  plot(
    pc1$x[, 1],
    pc1$x[, 2],
    col = as.numeric(gdp_factor),
    pch = 19,
    cex = 0.7,
    xlab = paste0("PC1 (", round(100 * prop_var[1], 1), "%)"),
    ylab = paste0("PC2 (", round(100 * prop_var[2], 1), "%)"),
    main = "PC1 vs PC2 Colored by GDP Group"
  )
  
  legend(
    "bottomright",
    legend = levels(gdp_factor),
    col = seq_along(levels(gdp_factor)),
    pch = 19,
    cex = 0.8,
    title = "GDP Group"
  )
}
```

---

## Step 14: PCA Scores and Loadings for Reporting

```{r}
pca_scores <- as.data.frame(pc1$x)
pca_scores$.row_id <- df_pca$.row_id

if ("microregion_code" %in% names(df_pca)) {
  pca_scores$microregion_code <- df_pca$microregion_code
}

if ("region" %in% names(df_pca)) {
  pca_scores$region <- df_pca$region
}

if ("gdp_group" %in% names(df_pca)) {
  pca_scores$gdp_group <- df_pca$gdp_group
}

metadata_cols <- intersect(
  c(".row_id", "microregion_code", "region", "gdp_group"),
  names(pca_scores)
)

pc_cols <- paste0("PC", seq_len(ncol(pc1$x)))
pca_scores <- pca_scores[, c(metadata_cols, pc_cols)]

# Print a compact score table for reporting.
pca_scores_compact <- pca_scores[
  ,
  c(metadata_cols, paste0("PC", 1:4))
]

head(pca_scores_compact)

pca_loadings_table <- data.frame(
  variable = rownames(pc1$rotation),
  pc1$rotation,
  row.names = NULL
)

# Print PC1-PC4 loadings, which are the retained components.
pca_loadings_pc1_pc4 <- pca_loadings_table[
  ,
  c("variable", paste0("PC", 1:4))
]

pca_loadings_pc1_pc4
```

---

## Final Summary

```{r}
cat("Final PCA variables:\n")
print(num_vars2)

cat("\nNumber of observations used in PCA:", nrow(df_pca), "\n")
cat("Number of variables used in PCA:", length(num_vars2), "\n")

cat("\nVariance explained by PC1:", round(100 * prop_var[1], 2), "%\n")
cat("Variance explained by PC2:", round(100 * prop_var[2], 2), "%\n")

pc12_text <- round(100 * cum_var[2], 2)
pc14_text <- round(100 * cum_var[min(4, length(cum_var))], 2)

cat("Cumulative variance explained by PC1-PC2:", pc12_text, "%\n")
cat("Cumulative variance explained by PC1-PC4:", pc14_text, "%\n")

n_kaiser <- sum(eigenvalues > 1)
cat("\nNumber of PCs retained by Kaiser criterion eigenvalue > 1:", n_kaiser, "\n")

cat("\nTop loadings for first four PCs:\n")
print(top_loadings)

cat("\nParallel analysis retention table:\n")
print(parallel_results)
```