---
title: "MANOVA"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MANOVA on Brazil Microregion Data

We test whether selected climate and environmental variables differ significantly across geographic regions. We also test whether selected climate and environmental variables differ among **Low**, **Medium**, and **High** GDP microregions.

For the GDP-group MANOVA, all three GDP groups are retained. **Low** is used as the reference category for the follow-up contrasts, so the pairwise contrasts are interpreted as:

- `Medium vs Low`
- `High vs Low`

Load required libraries.

```{r}
library(MASS)
library(biotools)

# Prevent rgl/X11 warnings when knitting on a server or Posit Cloud.
options(rgl.useNULL = TRUE)
library(heplots)

library(klaR)
library(car)
```

Load the prepared data and define the variables for analysis.

```{r}
load("/cloud/project/data/prepared_data.RData")

# Basic checks
stopifnot(exists("df"))
stopifnot(exists("num_vars"))

# Region MANOVA response variables.
# These are the variables used in the original region MANOVA analysis.
resp_vars_region <- c(
  "temp_mean", "temp_min", "temp_max",
  "humidity", "precip", "rainy_days", "ndvi"
)

# GDP-group MANOVA response variables.
# These match the original GDP-group MANOVA section.
resp_vars_gdp <- c(
  "temp_mean", "humidity", "precip", "rainy_days", "ndvi"
)

# Check that all required variables exist.
required_vars <- unique(c("region", "gdp_group", resp_vars_region, resp_vars_gdp))
missing_vars <- setdiff(required_vars, names(df))
if (length(missing_vars) > 0) {
  stop("The following required variables are missing from df: ",
       paste(missing_vars, collapse = ", "))
}

# Make sure region has a stable and interpretable order.
region_levels <- c("Centro-Oeste", "Nordeste", "Norte", "Sudeste", "Sul")
if (all(region_levels %in% unique(df$region))) {
  df$region <- factor(df$region, levels = region_levels)
} else {
  df$region <- factor(df$region)
}

# Make sure gdp_group has the intended order.
# Low is the reference group for follow-up treatment-contrast models.
df$gdp_group <- factor(df$gdp_group, levels = c("Low", "Medium", "High"))

# Use complete cases for each analysis.
df_region <- df[complete.cases(df[, c("region", resp_vars_region)]), ]
df_gdp <- df[complete.cases(df[, c("gdp_group", resp_vars_gdp)]), ]

# Confirm grouping variables.
table(df_region$region)
table(df_gdp$gdp_group)

# Confirm response variables.
resp_vars_region
resp_vars_gdp
```

---

## Step 1: Check Multivariate Normality Within Each Region

```{r}
# Chi-square Q-Q plots for each region.
for (i in seq_along(levels(df_region$region))) {
  current_region <- levels(df_region$region)[i]
  cqplot(df_region[df_region$region == current_region, resp_vars_region],
         main = paste(current_region))
}
```

---

## Step 2: Boxplots by Region

Visualize group differences for each response variable.

```{r}
for (v in resp_vars_region) {
  boxplot(df_region[[v]] ~ df_region$region,
          col = "lightblue",
          main = paste(v, "by Region"),
          xlab = "Region",
          ylab = v,
          cex.axis = 0.8)
  
  means <- tapply(df_region[[v]], df_region$region, mean, na.rm = TRUE)
  points(seq_along(means), means, col = "red", pch = 19, cex = 1.2)
  
  y_offset <- 0.04 * diff(range(df_region[[v]], na.rm = TRUE))
  text(x = seq_along(means),
       y = means + y_offset,
       labels = round(means, 2),
       cex = 0.8)
}
```

---

## Step 3: Descriptive Statistics by Region

```{r}
region_means <- aggregate(df_region[, resp_vars_region],
                          by = list(region = df_region$region),
                          FUN = mean)
region_means

region_sds <- aggregate(df_region[, resp_vars_region],
                        by = list(region = df_region$region),
                        FUN = sd)
region_sds
```

---

## Step 4: Check Homogeneity of Covariance Matrices by Region

Box's M test is sensitive to departures from normality, so it should be treated as a diagnostic rather than as a strict decision rule. When covariance matrices are heterogeneous, Pillai's trace is preferred as the primary MANOVA statistic.

```{r}
boxm_region <- try(
  biotools::boxM(df_region[, resp_vars_region], grouping = df_region$region),
  silent = TRUE
)
boxm_region
```

---

## Step 5: Run One-Way MANOVA by Region

```{r}
# Save the current contrast options so they can be restored at the end.
old_options <- options()

# Use sum contrasts for Type III tests.
options(contrasts = c("contr.sum", "contr.poly"))

manova_region <- lm(
  cbind(temp_mean, temp_min, temp_max,
        humidity, precip, rainy_days, ndvi) ~ region,
  data = df_region
)

# Univariate ANOVA results for each response variable.
summary.aov(manova_region)

# Multivariate tests.
anova(manova_region, test = "Pillai")
anova(manova_region, test = "Wilks")
anova(manova_region, test = "Hotelling-Lawley")
anova(manova_region, test = "Roy")
```

---

## Step 6: Combined Multivariate and Univariate Output by Region

```{r}
# Type III MANOVA using car::Anova.
region_anova_type3 <- car::Anova(manova_region, type = 3)

# Multivariate results only.
summary(region_anova_type3)

# Multivariate and univariate results together.
summary(region_anova_type3, univariate = TRUE)

# Multivariate and univariate results with Holm p-value adjustment.
summary(region_anova_type3, univariate = TRUE, p.adjust.method = "holm")
```

---

## Step 7: Check Residuals for Multivariate Normality

```{r}
cqplot(residuals(manova_region),
       main = "Residuals from Region MANOVA")
```

---

## Step 8: Correct Pairwise Contrasts Between Specific Regions

Important correction: with treatment contrasts, a coefficient such as `regionSul = 0` compares **Sul against the reference region**. Therefore, to test a specific pairwise comparison, we explicitly set the desired reference region before fitting the model.

```{r}
# Reset to treatment contrasts for direct pairwise interpretation.
options(contrasts = c("contr.treatment", "contr.poly"))

# -----------------------------
# 8.1 Sul vs Norte
# -----------------------------

# Set Norte as reference.
df_region_ref_norte <- df_region
df_region_ref_norte$region <- relevel(factor(df_region_ref_norte$region), ref = "Norte")

manova_region_ref_norte <- lm(
  cbind(temp_mean, temp_min, temp_max,
        humidity, precip, rainy_days, ndvi) ~ region,
  data = df_region_ref_norte
)

# With Norte as reference, regionSul = 0 tests Sul vs Norte.
contrast_sul_vs_norte <- car::linearHypothesis(
  manova_region_ref_norte,
  "regionSul = 0",
  test = "Pillai"
)
contrast_sul_vs_norte

# -----------------------------
# 8.2 Nordeste vs Norte
# -----------------------------

# With Norte as reference, regionNordeste = 0 tests Nordeste vs Norte.
contrast_nordeste_vs_norte <- car::linearHypothesis(
  manova_region_ref_norte,
  "regionNordeste = 0",
  test = "Pillai"
)
contrast_nordeste_vs_norte

# -----------------------------
# 8.3 Sul vs Sudeste
# -----------------------------

# Set Sudeste as reference.
df_region_ref_sudeste <- df_region
df_region_ref_sudeste$region <- relevel(factor(df_region_ref_sudeste$region), ref = "Sudeste")

manova_region_ref_sudeste <- lm(
  cbind(temp_mean, temp_min, temp_max,
        humidity, precip, rainy_days, ndvi) ~ region,
  data = df_region_ref_sudeste
)

# With Sudeste as reference, regionSul = 0 tests Sul vs Sudeste.
contrast_sul_vs_sudeste <- car::linearHypothesis(
  manova_region_ref_sudeste,
  "regionSul = 0",
  test = "Pillai"
)
contrast_sul_vs_sudeste
```

---

## Step 9: Univariate Follow-Up for Correct Region Contrasts

The multivariate contrasts above test whether the full response-variable profile differs between two regions. The following code provides univariate follow-up tests for the same selected pairwise region comparisons.

```{r}
# Function for univariate two-group comparison for selected region pairs.
region_pairwise_univariate <- function(data, region_a, region_b, response_vars) {
  subdat <- data[data$region %in% c(region_a, region_b), ]
  subdat$region <- factor(subdat$region, levels = c(region_a, region_b))
  
  out <- lapply(response_vars, function(v) {
    fit <- lm(as.formula(paste(v, "~ region")), data = subdat)
    a <- anova(fit)
    data.frame(
      comparison = paste(region_b, "vs", region_a),
      response = v,
      F_value = a$`F value`[1],
      p_value = a$`Pr(>F)`[1]
    )
  })
  
  out <- do.call(rbind, out)
  out$p_value_holm <- p.adjust(out$p_value, method = "holm")
  out
}

univ_sul_vs_norte <- region_pairwise_univariate(
  df_region,
  region_a = "Norte",
  region_b = "Sul",
  response_vars = resp_vars_region
)
univ_sul_vs_norte

univ_nordeste_vs_norte <- region_pairwise_univariate(
  df_region,
  region_a = "Norte",
  region_b = "Nordeste",
  response_vars = resp_vars_region
)
univ_nordeste_vs_norte

univ_sul_vs_sudeste <- region_pairwise_univariate(
  df_region,
  region_a = "Sudeste",
  region_b = "Sul",
  response_vars = resp_vars_region
)
univ_sul_vs_sudeste
```

---

## Step 10: MANOVA by GDP Group - Low / Medium / High

We test whether selected climate and environmental variables differ among **Low**, **Medium**, and **High** GDP microregions. Unlike the earlier High-vs-Medium binary analysis, this section retains all three GDP groups. Low is used as the reference level for follow-up contrasts.

```{r}
# Keep all three GDP groups.
df_gdp_all <- df_gdp[df_gdp$gdp_group %in% c("Low", "Medium", "High"), ]
df_gdp_all$gdp_group <- factor(df_gdp_all$gdp_group,
                               levels = c("Low", "Medium", "High"))

# Confirm group sizes.
table(df_gdp_all$gdp_group)
levels(df_gdp_all$gdp_group)

# Descriptive statistics by GDP group.
gdp_group_means <- aggregate(df_gdp_all[, resp_vars_gdp],
                             by = list(gdp_group = df_gdp_all$gdp_group),
                             FUN = mean)
gdp_group_means

gdp_group_sds <- aggregate(df_gdp_all[, resp_vars_gdp],
                           by = list(gdp_group = df_gdp_all$gdp_group),
                           FUN = sd)
gdp_group_sds
```

---

## Step 11: Boxplots by GDP Group

```{r}
for (v in resp_vars_gdp) {
  boxplot(df_gdp_all[[v]] ~ df_gdp_all$gdp_group,
          col = "lightblue",
          main = paste(v, "by GDP Group"),
          xlab = "GDP Group",
          ylab = v)
  
  means <- tapply(df_gdp_all[[v]], df_gdp_all$gdp_group, mean, na.rm = TRUE)
  points(seq_along(means), means, col = "red", pch = 19, cex = 1.2)
  
  y_offset <- 0.04 * diff(range(df_gdp_all[[v]], na.rm = TRUE))
  text(x = seq_along(means),
       y = means + y_offset,
       labels = round(means, 2),
       cex = 0.8)
}
```

---

## Step 12: Check Multivariate Normality and Covariance Homogeneity by GDP Group

```{r}
for (i in seq_along(levels(df_gdp_all$gdp_group))) {
  current_group <- levels(df_gdp_all$gdp_group)[i]
  cqplot(df_gdp_all[df_gdp_all$gdp_group == current_group, resp_vars_gdp],
         main = paste("MVN Check - GDP Group", current_group))
}

boxm_gdp_all <- try(
  biotools::boxM(df_gdp_all[, resp_vars_gdp], grouping = df_gdp_all$gdp_group),
  silent = TRUE
)
boxm_gdp_all
```

---

## Step 13: Run MANOVA by GDP Group - Overall Low / Medium / High Effect

```{r}
# Use sum contrasts for Type III tests of the overall GDP-group effect.
options(contrasts = c("contr.sum", "contr.poly"))

manova_gdp_all <- lm(
  cbind(temp_mean, humidity, precip, rainy_days, ndvi) ~ gdp_group,
  data = df_gdp_all
)

# Univariate ANOVA results for each response variable.
summary.aov(manova_gdp_all)

# Multivariate tests.
anova(manova_gdp_all, test = "Pillai")
anova(manova_gdp_all, test = "Wilks")
anova(manova_gdp_all, test = "Hotelling-Lawley")
anova(manova_gdp_all, test = "Roy")

# Type III MANOVA using car::Anova.
gdp_all_anova_type3 <- car::Anova(manova_gdp_all, type = 3)
summary(gdp_all_anova_type3)
summary(gdp_all_anova_type3, univariate = TRUE)
summary(gdp_all_anova_type3, univariate = TRUE, p.adjust.method = "holm")
```

---

## Step 14: GDP Group Contrasts with Low as Reference

This step directly addresses the target interpretation: **Medium vs Low** and **High vs Low**. With treatment contrasts and `Low` as the reference level:

- `gdp_groupMedium = 0` tests `Medium vs Low`
- `gdp_groupHigh = 0` tests `High vs Low`

```{r}
# Use treatment contrasts for direct Low-reference interpretation.
options(contrasts = c("contr.treatment", "contr.poly"))

df_gdp_ref_low <- df_gdp_all
df_gdp_ref_low$gdp_group <- relevel(factor(df_gdp_ref_low$gdp_group), ref = "Low")

# Confirm that Low is the reference category.
levels(df_gdp_ref_low$gdp_group)
contrasts(df_gdp_ref_low$gdp_group)

manova_gdp_ref_low <- lm(
  cbind(temp_mean, humidity, precip, rainy_days, ndvi) ~ gdp_group,
  data = df_gdp_ref_low
)

# Medium vs Low multivariate contrast.
contrast_medium_vs_low <- car::linearHypothesis(
  manova_gdp_ref_low,
  "gdp_groupMedium = 0",
  test = "Pillai"
)
contrast_medium_vs_low

# High vs Low multivariate contrast.
contrast_high_vs_low <- car::linearHypothesis(
  manova_gdp_ref_low,
  "gdp_groupHigh = 0",
  test = "Pillai"
)
contrast_high_vs_low

# Joint test: Medium vs Low and High vs Low simultaneously.
# This is equivalent to testing whether the GDP-group profile differs from Low for either non-Low group.
contrast_medium_high_vs_low_joint <- car::linearHypothesis(
  manova_gdp_ref_low,
  c("gdp_groupMedium = 0", "gdp_groupHigh = 0"),
  test = "Pillai"
)
contrast_medium_high_vs_low_joint
```

---

## Step 15: Univariate Follow-Up for GDP Group Contrasts with Low as Reference

The multivariate tests above evaluate the full response-variable profile. The table below gives univariate follow-up results for `Medium vs Low` and `High vs Low` for each response variable. Holm adjustment is applied across all follow-up p-values.

```{r}
gdp_low_ref_univariate <- lapply(resp_vars_gdp, function(v) {
  fit <- lm(as.formula(paste(v, "~ gdp_group")), data = df_gdp_ref_low)
  coef_tab <- summary(fit)$coefficients
  
  data.frame(
    response = v,
    comparison = c("Medium vs Low", "High vs Low"),
    estimate_difference = c(coef_tab["gdp_groupMedium", "Estimate"],
                            coef_tab["gdp_groupHigh", "Estimate"]),
    std_error = c(coef_tab["gdp_groupMedium", "Std. Error"],
                  coef_tab["gdp_groupHigh", "Std. Error"]),
    t_value = c(coef_tab["gdp_groupMedium", "t value"],
                coef_tab["gdp_groupHigh", "t value"]),
    p_value = c(coef_tab["gdp_groupMedium", "Pr(>|t|)"],
                coef_tab["gdp_groupHigh", "Pr(>|t|)"])
  )
})

gdp_low_ref_univariate <- do.call(rbind, gdp_low_ref_univariate)
gdp_low_ref_univariate$p_value_holm <- p.adjust(gdp_low_ref_univariate$p_value,
                                                method = "holm")
gdp_low_ref_univariate
```

---

## Step 16: Check Residuals for GDP Group MANOVA

```{r}
cqplot(residuals(manova_gdp_all),
       main = "Residuals from GDP Group MANOVA: Low / Medium / High")
```

---

## Step 17: Final Summary

```{r}
cat("Region MANOVA response variables:", paste(resp_vars_region, collapse = ", "), "\n")
cat("GDP-group MANOVA response variables:", paste(resp_vars_gdp, collapse = ", "), "\n")
cat("GDP-group levels:", paste(levels(df_gdp_all$gdp_group), collapse = ", "), "\n")
cat("GDP-group follow-up reference level: Low\n")
cat("GDP-group follow-up contrasts: Medium vs Low; High vs Low\n")
```

---

## Step 18: Restore Original Options

```{r}
# Restore original global options.
options(old_options)
```